---
title: "Introduction to Research Data Analysis"
subtitle: "From Research Question to Defensible Analysis — A Workshop for Part II Dissertation Candidates (NPMCN / WACP)"
author: "Stephen Balogun (RStudio [now Posit] Certified tidyverse and shiny instructor"
date: today
format:
  revealjs:
    embed-resources: true
    theme: [default, custom.scss]
    slide-number: true
    progress: true
    controls: true
    hash: true
    transition: fade
    incremental: false
    center: false
    code-fold: true
    code-tools: true
    chalkboard: true
    preview-links: auto
    width: 1280
    height: 720
    margin: 0.06
    logo: images/logo.png
    footer: "Introduction to Research Data Analysis | NPMCN/WACP Part II Prep"
    include-after-body: scripts/interactions.html
execute:
  echo: true
  warning: false
  message: false
bibliography: references.bib
---

```{r setup}
#| include: false
library(tidyverse)
library(gt)
library(scales)

set.seed(2026)

n <- 742

hypertension_demo <- tibble(
  participant_id = sprintf("HUC-%04d", 1:n),
  age = round(rnorm(n, mean = 42, sd = 13)) |> pmax(18) |> pmin(85),
  sex = sample(c("Male", "Female"), n, replace = TRUE, prob = c(0.47, 0.53)),
  lga = sample(paste("LGA", 1:6), n, replace = TRUE),
  education = sample(
    c("None", "Primary", "Secondary", "Tertiary"),
    n, replace = TRUE, prob = c(0.10, 0.25, 0.40, 0.25)
  ),
  smoking = rbinom(n, 1, 0.18),
  physical_activity = sample(
    c("Low", "Moderate", "High"), n, replace = TRUE, prob = c(0.35, 0.45, 0.20)
  ),
  bmi = round(rnorm(n, mean = 26, sd = 4.5), 1) |> pmax(15) |> pmin(48),
  monthly_income = round(rlnorm(n, meanlog = 11, sdlog = 0.6))
) |>
  mutate(
    # Build correlated risk: age, bmi, smoking increase hypertension log-odds
    logit_htn = -6.5 + 0.05 * age + 0.09 * bmi + 0.6 * smoking,
    p_htn = plogis(logit_htn),
    hypertension = rbinom(n, 1, p_htn),
    systolic_bp = round(100 + 0.6 * age + 0.5 * bmi + 8 * smoking + rnorm(n, 0, 9)),
    diastolic_bp = round(60 + 0.25 * age + 0.3 * bmi + rnorm(n, 0, 7)),
    diabetes = rbinom(n, 1, plogis(-5 + 0.04 * age + 0.07 * bmi)),
    medication_use = if_else(hypertension == 1, rbinom(n, 1, 0.55), 0),
    hypertension = factor(hypertension, labels = c("No", "Yes")),
    diabetes = factor(diabetes, labels = c("No", "Yes")),
    smoking = factor(smoking, labels = c("Non-smoker", "Smoker"))
  ) |>
  select(-logit_htn, -p_htn)

# write for reuse / participant download
write_csv(hypertension_demo, "data/hypertension_demo.csv")

pal <- list(navy = "#0B2545", teal = "#1B7F79", green = "#5B8C5A",
            amber = "#E0A458", bg = "#F7F5F0", ink = "#1C1C1C")

theme_workshop <- function() {
  theme_minimal(base_size = 16) +
    theme(
      plot.title = element_text(face = "bold", color = pal$navy, size = 18),
      plot.subtitle = element_text(color = "grey30"),
      panel.grid.minor = element_blank(),
      axis.title = element_text(color = "grey20")
    )
}
```

::: notes
Simulated "Hypertension in Urban Communities Study" (n = 742) is the running case used across all modules. All 742 rows are simulated — state this explicitly to participants at first use (Module 1, slide "Meet the dataset").
:::

# Module 1 — Welcome, Research Foundations & the Analysis Journey {background-color="#0B2545"}

## You've collected your data. Now what? {.center}

::: r-fit-text
**742 rows.**
:::

You have completed data collection for your dissertation.

What happens *next* — and why does it matter more than which software you open?

::: notes
Opening hook. Ask verbally: "How many of you already have a dataset sitting on your laptop right now?" Transition into objectives.
:::

------------------------------------------------------------------------

### Workshop Expectations

- 120 minutes, 7 short modules — we move briskly

- You will be asked to **commit to an answer** before we reveal it (chat / poll / reaction)

- One running case study throughout: the *Hypertension in Urban Communities Study*

- Goal is not to make you a statistician — it is to make you a **defensible decision-maker**

::: notes
Set norms for Teams delivery: chat responses, reactions, polls. Mention breakout rooms if available.
:::

------------------------------------------------------------------------

### Learning Objectives {.smaller}

By the end of this workshop you will be able to:

::::: columns
::: {.column width="50%"}
1.  Link objective → design → variables → analysis
2.  Review common PH study designs
3.  Revise sample-size/sampling logic
4.  Classify variables (categorical/continuous, NOIR)
5.  Spot common data-quality problems
6.  Choose descriptive statistics & visuals
:::

::: {.column width="50%"}
7.  Distinguish descriptive vs inferential analysis
8.  Interpret p-values & 95% CIs correctly
9.  Select the right statistical test
10. Explain *why* regression is needed
11. Interpret linear & logistic regression output
12. Build a defensible analysis plan
:::
:::::

::: notes
Don't read all 12 aloud — point out this slide will resurface as a "recall card" at the very end.
:::

------------------------------------------------------------------------

### 🗳️ Opening Diagnostic Poll {background-color="#1B7F79"}

**POLL (Teams):** *How confident are you right now in choosing the correct statistical test for your dissertation objective?*

- A. Very confident
- B. Somewhat confident
- C. Not confident / guessing

::: notes
Launch a live Teams poll. Do not comment on results yet — revisit at closing to show growth. No right answer; this is a baseline, not a quiz.
:::

------------------------------------------------------------------------

### The Analysis Journey — Our Map for Today

```{mermaid}
flowchart LR
  A[Research<br>Question] --> B[Study<br>Design]
  B --> C[Variables]
  C --> D[Data<br>Quality]
  D --> E[Descriptive<br>Analysis]
  E --> F[Bivariate<br>Analysis]
  F --> G[Regression]
  G --> H[Interpretation<br>& Reporting]
```

::: fragment
**🧭 Where are we?** This banner returns at the start of every module.
:::

::: notes
This is the recurring "Where are we in the analysis journey?" visual — screenshot or rebuild a simplified version to insert at the top of every module title slide.
:::

------------------------------------------------------------------------

### Core Takeaway #1 {.center background-color="#0B2545"}

::: r-fit-text
Analysis begins with the **research question** and **study design** —

not with opening SPSS, Stata, or R.
:::

------------------------------------------------------------------------

### Rapid Review — Study Design Landscape

```{mermaid}
flowchart TD
  S[Study Designs] --> O[Observational]
  S --> E[Experimental]
  O --> D1[Descriptive<br>Cross-sectional]
  O --> D2[Analytical]
  D2 --> CC[Case-control]
  D2 --> CO[Cohort]
  D2 --> CS[Analytical<br>Cross-sectional]
  E --> RCT[RCT /<br>Community Trial]
```

::: notes
Emphasize: descriptive cross-sectional studies describe "what/who/where"; analytical designs test "why/is there an association".
:::

------------------------------------------------------------------------

### Design → Question It Answers

| Design | Typical Question | Direction of Inquiry | Key Output |
|------------------|------------------|------------------|------------------|
| Cross-sectional (descriptive) | How common is X? | Snapshot | Prevalence |
| Case-control | What's associated with disease? | Outcome → exposure | Odds Ratio |
| Cohort | Does exposure predict outcome? | Exposure → outcome | Risk Ratio / Incidence |
| RCT / Community trial | Does intervention change outcome? | Allocated exposure → outcome | Risk difference / RR |

::: notes
Ask participants (chat): "Which of these is your dissertation design?" Use responses to tailor emphasis for the rest of the workshop.
:::

------------------------------------------------------------------------

### ✅ Identify-the-Design Check

A resident recruits 300 hypertensive and 300 normotensive adults from a clinic and looks *backward* at their salt-intake history.

::: {.fragment .fade-in}
**A.** Cohort   **B.** Case-control   **C.** RCT   **D.** Cross-sectional
:::

::: {.fragment .fade-in}
**Answer: B — Case-control.** Outcome status (hypertensive vs not) is fixed at recruitment; exposure (salt intake) is measured retrospectively.
:::

::: notes
Commit-reveal-explain. Ask for chat answers before advancing the fragment. Common wrong answer: cross-sectional — clarify that the *backward-looking, outcome-anchored sampling* is the case-control signature.
:::

------------------------------------------------------------------------

### Bridge: Design → Variables → Analysis

::: r-fit-text
Your **study design** constrains which **variables** you can measure —

which constrains which **statistics** are valid.
:::

Get this chain wrong, and no software will save you. Next: sample size and sampling.

# Module 2 — Sample Size & Sampling {background-color="#0B2545"}

## 🧭 Where are we? — Design ✅ → **Sample Size** → Variables → ...

------------------------------------------------------------------------

### Why Sample Size Matters

- Too small → underpowered, real effects missed (Type II error)
- Too large → wasted resources, trivial effects become "significant"
- Examiners *will* ask: "How did you arrive at this number?"
- A justified sample size = a defensible Chapter 3

------------------------------------------------------------------------

### Single-Proportion Sample Size (Cochran)

$$n = \frac{Z^2 \, P(1-P)}{d^2}$$

| Symbol | Meaning                                                   |
|--------|-----------------------------------------------------------|
| $Z$    | Standard normal value for confidence level (1.96 for 95%) |
| $P$    | Expected proportion (from prior literature/pilot)         |
| $d$    | Margin of error (precision) you will accept               |

::: fragment
We are **not** deriving this — we are learning what each term forces you to *justify*.
:::

------------------------------------------------------------------------

### Worked Example

> Expected hypertension prevalence in an urban LGA = **30%** (from prior study). You want **5%** precision at 95% confidence.

```{r sample-size-example}
Z <- 1.96
P <- 0.30
d <- 0.05

n_required <- (Z^2 * P * (1 - P)) / d^2
ceiling(n_required)
```

::: fragment
**≈ 323 participants**, before adjustments.
:::

::: notes
Walk through substitution live. Ask: "What happens to n if P were 50% instead of 30%?" (Answer: n increases — variance is maximized at P = 0.5.)
:::

------------------------------------------------------------------------

### Adjustments You Will Likely Need

::::: columns
::: {.column width="50%"}
**Finite population correction** $$n_{adj} = \frac{n}{1 + (n-1)/N}$$ Used when sampling frame (N) is small/known.

**Non-response adjustment** $$n_{final} = \frac{n}{1 - \text{(expected non-response \%)}}$$
:::

::: {.column width="50%"}
**Design effect (DEFF)** $$n_{cluster} = n \times DEFF$$ Applied when sampling is **not** simple random (e.g., cluster/multistage) — accounts for reduced precision from clustering.
:::
:::::

::: notes
Emphasize: DEFF is often the most forgotten adjustment in dissertations that use multistage sampling but calculate n as if SRS.
:::

------------------------------------------------------------------------

### Comparing Two Groups (Conceptual)

For comparing **two proportions** or **two means**, sample size formulas add:

- Expected difference you want to be able to detect (effect size)
- Variability in each group
- Desired **power** (usually 80–90%) in addition to confidence level
- Larger detectable difference → smaller required sample

::: fragment
You will rarely hand-calculate these — but you must be able to **defend every input** you fed into G\*Power/Stata/R.
:::

------------------------------------------------------------------------

### Sampling Methods at a Glance

```{mermaid}
flowchart TD
  P[Sampling Methods] --> Prob[Probability]
  P --> NonProb[Non-probability]
  Prob --> SRS[Simple Random]
  Prob --> Sys[Systematic]
  Prob --> Strat[Stratified]
  Prob --> Multi[Multistage / Cluster]
  NonProb --> Purp[Purposive]
  NonProb --> Snow[Snowball]
```

::: fragment
⚠️ Non-probability sampling **limits generalizability** — say so explicitly in your limitations.
:::

::: notes
Ask participants which method their proposed dissertation uses; multistage/cluster is common for community-based PH surveys in Nigeria.
:::

------------------------------------------------------------------------

### ✅ Quick Check — Sampling

A resident selects every 5th patient file from the clinic register after a random start point. This is:

::: fragment
**A.** Stratified   **B.** Systematic   **C.** Cluster   **D.** Purposive
:::

::: fragment
**Answer: B — Systematic sampling.** Fixed interval + random start.
:::

# Module 3 — Variables & Measurement Scales {background-color="#0B2545"}

## 🧭 Where are we? — Design ✅ → Sample Size ✅ → **Variables** → ...

------------------------------------------------------------------------

### Two Fundamental Roles

::::: columns
::: {.column width="50%"}
#### 🎯 Outcome / Dependent

The thing you are trying to explain or predict.

*e.g., hypertension status, systolic BP*
:::

::: {.column width="50%"}
#### 🔧 Exposure / Predictor / Independent

The thing you think influences the outcome.

*e.g., smoking, age, BMI*
:::
:::::

::: fragment
Every dissertation objective should let you point at both.
:::

------------------------------------------------------------------------

### Continuous vs Discrete vs Categorical

| Type        | Definition              | Example                 |
|-------------|-------------------------|-------------------------|
| Continuous  | Any value on a range    | Age, BMI, SBP           |
| Discrete    | Countable whole numbers | Number of pregnancies   |
| Categorical | Named groups            | Sex, LGA, disease stage |

Variables can be converted from one form to another

```{mermaid}
flowchart LR
  C[Continuous] --> D[Discrete]
  D --> K[Categorical]
```

------------------------------------------------------------------------

### NOIR — The Measurement Scale Ladder

```{mermaid}
flowchart LR
  N[Nominal<br>named, unordered] --> O[Ordinal<br>ordered, unequal gaps]
  O --> I[Interval<br>ordered, equal gaps,<br>no true zero]
  I --> R[Ratio<br>ordered, equal gaps,<br>true zero]
```

| Scale | Key idea | Public-health examples |
|-------------------|---------------------|--------------------------------|
| **Nominal** | Categories with **no inherent order** | Hypertension status (yes/no); vaccination status (vaccinated/unvaccinated); blood group (A/B/AB/O); LGA of residence; marital status; disease diagnosis; type of health facility |
| **Ordinal** | Categories have a **meaningful order**, but gaps are not necessarily equal | Cancer stage (I–IV); disease severity (mild/moderate/severe); educational attainment; socioeconomic class; Likert response (strongly disagree → strongly agree); nutritional status categories |
| **Interval** | Equal intervals, but **zero is arbitrary** | Temperature in °C; calendar year; date of diagnosis/date of birth |
| **Ratio** | Equal intervals **and a meaningful true zero** | Age; weight; height; systolic BP; CD4 count; monthly income; duration of illness; number of clinic visits |

::: notes
Common confusion: temperature in °C is interval (no true zero — 0°C isn't "no temperature"); Kelvin would be ratio. Age/BMI/income all have a true zero and are ratio.
:::

------------------------------------------------------------------------

### Tabular Data

Data is collected into a form of spreadsheet with rows and variables

Rows: horizontal (represents all measurements from a single unique observation)

Columns: vertical (a single attribute measured across all observations)

::: fragment
**📌 The Golden Rule of Tidy Data**

To keep data "tidy" and ready for analysis, every dataset must follow this strict structure:

Each variable must have its own column.

Each observation must have its own row.

Each value (individual cell) must contain a single measurement.
:::

------------------------------------------------------------------------

### 🎯 Classify-the-Variable Round (Quiz)

::: fragment
**1. Which of the followings is not a categorical variable?**

A)  Place of delivery

B)  Occupation

C)  Salary

D)  Ethnicity
:::

::: fragment
**2. Which of the followings is a continuous variable?**

A)  Heart rate

B)  Eye Colour

C)  CGPA
:::

::: fragment
**3. Which of the variable(s) below does not follow the "tidy data" principle?**

A)  Blood Pressure

B)  BMI

C)  Heart rate
:::

::: notes
Go one at a time; ask chat for answer before revealing each fragment. The Likert item is the deliberate trap — many residents wrongly average Likert scores as if interval.
:::

------------------------------------------------------------------------

### ⚠️ Why This Matters: The Bridge Forward

::: r-fit-text
**Variable Type → Summary Measure → Visualization → Statistical Test**
:::

Get the variable type wrong, and every downstream choice — the table you build, the graph you draw, the test you run — is built on sand.

::: notes
This exact phrase should be echoed again at the start of Module 5. Consider a persistent footer icon or recurring slide graphic.
:::

------------------------------------------------------------------------

### ✅ Choose the Summary — Preview

`hypertension` (Yes/No) should be summarized as:

::: fragment
**A.** Mean ± SD   **B.** Frequency & percentage   **C.** Median (IQR)
:::

::: fragment
**Answer: B.** Categorical variables are summarized with counts/percentages, never means.
:::

# Module 4 — Data Collection, Management & Quality Assurance {background-color="#0B2545"}

## 🧭 Where are we? — ... → Variables ✅ → **Data Quality** → ...

------------------------------------------------------------------------

### Tools You'll Likely Encounter

::::: columns
::: {.column width="50%"}
**Digital**

- KoboToolbox
- ODK
- REDCap
:::

::: {.column width="50%"}
**Paper-based**

- Interviewer-administered questionnaires
- Checklists
:::
:::::

::: fragment
Digital tools reduce (but do not eliminate) data-entry error — validation rules must still be built in deliberately.
:::

------------------------------------------------------------------------

### From Raw Data to Analysis-Ready

```{mermaid}
flowchart LR
  Raw[Raw Data] --> Val[Validation]
  Val --> Clean[Cleaning]
  Clean --> Recode[Recoding /<br>Derived Variables]
  Recode --> Ready[Analysis-Ready<br>Dataset]
```

::: fragment
**Data cleaning is analysis** — not an administrative chore before "the real work."
:::

------------------------------------------------------------------------

### 🔍 Spot the Problem — 30-Second Challenge

| id  | age     | sex   | sbp     | dob        | hypertension |
|-----|---------|-------|---------|------------|--------------|
| 001 | 34      | M     | 128     | 1990-04-12 | No           |
| 002 | **250** | F     | 118     | 1985-01-30 | Yes          |
| 003 | 45      | **1** | **-20** | 2031-06-01 | Yes          |
| 001 | 34      | M     | 128     | 1990-04-12 | No           |

::: fragment
**Errors:** 1. implausible age (250)

2.  mixed sex coding (M/F vs 1/2)

3.  negative SBP

4.  future/impossible date

5.  **duplicate ID 001**
:::

::: notes
Give participants 30 real seconds via chat before revealing. This mirrors the required "messy mini-dataset" exercise.
:::

------------------------------------------------------------------------

### Common Data-Quality Problems Checklist

- Missing values (and *why* they're missing — random vs systematic)
- Duplicate records/IDs
- Impossible or out-of-range values
- Inconsistent categorical coding across data collectors
- Outliers — real vs data-entry error
- BMI inconsistent with recorded height/weight
- Unnecessary categorization of continuous variables (loses information/power)

::: notes
Stress the "unnecessary categorization" point — a frequent dissertation error is converting a perfectly good continuous variable into categories without justification, discarding statistical power.
:::

------------------------------------------------------------------------

### Ethics Reminder

::: r-fit-text
Anonymity. Confidentiality. Secure storage.

Every dataset in this workshop is **simulated** — none of these 742 "participants" exist.
:::

# Module 5 — Descriptive & Inferential Statistics {background-color="#0B2545"}

## 🧭 Where are we? — ... → Data Quality ✅ → **Descriptive & Inferential** → ...

------------------------------------------------------------------------

## Part A: Descriptive Statistics

### Why Describe Before You Test?

- You cannot interpret a p-value without knowing what "normal" looks like in your data
- Descriptive statistics reveal data-quality problems you missed earlier
- Every dissertation needs a **Table 1** — this is where it comes from

------------------------------------------------------------------------

### Measures of Central Tendency and Measures of Dispersion

::::: columns
::: {.column width="50%"}
**Measures of Central Tendency**

- Mean
- Median
- Mode
:::

::: {.column width="50%"}
**Measures of Dispersion**

- Standard Deviation
- Inter Quartile Range
- Variance
- Range
:::
:::::

------------------------------------------------------------------------

### Distribution Shapes

```{r distributions}
#| fig-height: 4
set.seed(1)
df_shapes <- tibble(
  symmetric = rnorm(2000, 50, 10),
  right_skew = rgamma(2000, shape = 2, scale = 10),
  left_skew = 100 - rgamma(2000, shape = 2, scale = 10)
) |> pivot_longer(everything(), names_to = "shape", values_to = "value")

ggplot(df_shapes, aes(value, fill = shape)) +
  geom_histogram(bins = 40, show.legend = FALSE) +
  facet_wrap(~shape, scales = "free") +
  scale_fill_manual(values = c(pal$teal, pal$amber, pal$green)) +
  theme_workshop() +
  labs(title = "Symmetric vs Skewed Distributions", x = NULL, y = NULL)
```

::: notes
Point out: symmetric → mean ≈ median; right-skew → mean pulled above median; left-skew → mean pulled below median.
:::

------------------------------------------------------------------------

### Mean vs Median: The Outlier Effect

Generally, there are tests to check for skewness (Normality tests)

Quick "Eyeball" methods include:

1.  Histogram

2.  Boxplot

3.  Q-Q plot

::: fragment
**Rule of thumb:** skewed data → report **median (IQR)**, not mean ± SD.
:::

------------------------------------------------------------------------

### Choosing Mean±SD vs Median(IQR)

```{r sbp-check}
hypertension_demo |>
  summarise(
    mean_sbp = mean(systolic_bp), sd_sbp = sd(systolic_bp),
    median_sbp = median(systolic_bp),
    q1 = quantile(systolic_bp, 0.25), q3 = quantile(systolic_bp, 0.75)
  ) |>
  gt() |> fmt_number(decimals = 1)
```

::: fragment
Mean and median are close here → SBP is approximately symmetric → **mean ± SD is defensible**.
:::

------------------------------------------------------------------------

### Visualization Gallery — Match the Graph to the Variable

:::::: columns
::: {.column width="33%"}
**Categorical** Bar chart
:::

::: {.column width="33%"}
**Continuous (shape)** Histogram
:::

::: {.column width="33%"}
**Continuous (comparison)** Box plot
:::
::::::

```{r viz-gallery}
#| fig-height: 3.2
library(patchwork)

p1 <- ggplot(hypertension_demo, aes(hypertension, fill = hypertension)) +
  geom_bar() + scale_fill_manual(values = c(pal$teal, pal$amber)) +
  theme_workshop() + theme(legend.position = "none") + labs(title = "Bar Chart", x=NULL,y=NULL)

p2 <- ggplot(hypertension_demo, aes(systolic_bp)) +
  geom_histogram(bins = 30, fill = pal$navy) +
  theme_workshop() + labs(title = "Histogram", x=NULL,y=NULL)

p3 <- ggplot(hypertension_demo, aes(hypertension, systolic_bp, fill = hypertension)) +
  geom_boxplot(show.legend = FALSE) + scale_fill_manual(values = c(pal$teal, pal$amber)) +
  theme_workshop() + labs(title = "Box Plot", x=NULL,y=NULL)

p1 | p2 | p3
```

::: fragment
https://stbalo2002.shinyapps.io/Intro_to_biostatistics
:::

------------------------------------------------------------------------

### ✅ Choose the Graph

You want to show the **relationship between age and systolic BP** (both continuous). Best graph?

::: fragment
**A.** Bar chart   **B.** Box plot   **C.** Scatterplot   **D.** Pie chart
:::

::: fragment
**Answer: C — Scatterplot.** Two continuous variables → scatterplot (sets up Module 6's regression line).
:::

------------------------------------------------------------------------

## Part B: Inferential Statistics

### Population vs Sample

::: r-fit-text
We never measure the **population**.

We measure a **sample** — and use it to make a careful, uncertain statement about the population.
:::

------------------------------------------------------------------------

### What a P-value Actually Is

::: fragment
❌ *"The probability the null hypothesis is true."*
:::

::: fragment
✅ **The p-value quantifies how incompatible the observed data are with the null model, under the assumptions of the test.**
:::

::: fragment
In plain terms: *"If there really were no effect, how surprising would data like ours be?"*
:::

::: notes
This is the single most important misconception to correct all workshop. Slow down here.
:::

------------------------------------------------------------------------

### The 95% Confidence Interval

::: incremental
- A **range of plausible values** for the true population effect, given your sample
- "95% confidence" describes the **procedure's long-run reliability**, not this specific interval
- A CI that **excludes the null value** (0 for differences, 1 for ratios) aligns with statistical significance
- **Width tells you precision** — wide CI = imprecise, even if "significant"
:::

------------------------------------------------------------------------

### Statistical Power & Errors (Introductory)

|                       | H0 True          | H0 False          |
|-----------------------|------------------|-------------------|
| **Reject H0**         | Type I error (α) | Correct decision  |
| **Fail to reject H0** | Correct decision | Type II error (β) |

::: fragment
**Power = 1 − β** → probability of detecting a real effect when one exists. This is why Module 2's sample size mattered.
:::

------------------------------------------------------------------------

### The Reusable Decision Sequence

::: incremental
1.  What is my research objective?
2.  What is my outcome variable — and its type?
3.  What is my explanatory/grouping variable — and its type?
4.  How many groups or measurements?
5.  Are observations independent or paired?
6.  Is the relevant distribution compatible with parametric assumptions?
7.  What effect estimate/test actually answers the question?
:::

::: fragment
Memorize this **sequence**, not a giant table of tests.
:::

------------------------------------------------------------------------

### 🧰 Which Statistical Test Should I Use: The Interactive Decision Pathway

```{mermaid}
flowchart TD
  Obj[Objective] --> Desc[Describe]
  Obj --> Comp[Compare]
  Obj --> Assoc[Associate]
  Obj --> Pred[Predict]
  Comp --> OutType{Outcome type?}
  OutType -->|Categorical| Cat[Chi-square /<br>Fisher's exact]
  OutType -->|Continuous| Groups{How many<br>groups?}
  Groups -->|2, independent| T1[t-test /<br>Mann-Whitney U]
  Groups -->|2, paired| T2[Paired t-test /<br>Wilcoxon signed-rank]
  Groups -->|3+| T3[ANOVA /<br>Kruskal-Wallis]
  Assoc --> Corr[Pearson /<br>Spearman correlation]
  Pred --> Reg[Linear /<br>Logistic regression]
```

::: fragment
⚠️ **A decision tree is a starting aid — not a substitute for understanding your design, assumptions, and question.**
:::

::: notes
This is the flagship interactive tool referenced in the brief. Walk through it live using the running case: "Is hypertension associated with smoking?" → categorical outcome, categorical exposure → chi-square.
:::

------------------------------------------------------------------------

### Test Card: Chi-square / Fisher's Exact

|                |                                                            |
|------------------------------------|------------------------------------|
| **Question**   | Is there an association between two categorical variables? |
| **Variables**  | Categorical outcome & categorical exposure                 |
| **Assumption** | Expected cell counts ≥5 (else use Fisher's exact)          |
| **PH Example** | Hypertension (Y/N) vs Smoking (Y/N)                        |
| **Report as**  | χ²(df) = value, p = value, or OR with 95% CI               |

```{r chisq-demo}
tab <- table(hypertension_demo$smoking, hypertension_demo$hypertension)
chisq.test(tab)
```

------------------------------------------------------------------------

### ✅ Choose the Test

A researcher compares mean SBP between smokers and non-smokers. SBP is approximately normal; groups are independent.

::: fragment
**A.** Chi-square   **B.** Independent t-test   **C.** Paired t-test   **D.** Spearman correlation
:::

::: fragment
**Answer: B — Independent t-test.** Continuous outcome, two independent groups, approximately normal.
:::

```{r ttest-demo}
#| eval: false
t.test(systolic_bp ~ smoking, data = hypertension_demo)
```

------------------------------------------------------------------------

### Test Card: Correlation

|  |  |
|------------------------------------|------------------------------------|
| **Pearson** | Linear association between two continuous, roughly normal variables |
| **Spearman** | Monotonic association; ordinal data or non-normal continuous data |
| **PH Example** | Age vs systolic BP |
| **Reminder** | Correlation ≠ causation; check linearity visually first |

```{r scatter-demo}
#| fig-height: 3.5
ggplot(hypertension_demo, aes(age, systolic_bp)) +
  geom_point(alpha = 0.3, color = pal$teal) +
  geom_smooth(method = "lm", color = pal$navy, se = TRUE) +
  theme_workshop() +
  labs(title = "Age vs Systolic BP", x = "Age (years)", y = "SBP (mmHg)")
```

------------------------------------------------------------------------

### ✅ Choose the Test — Round 2

Compare hypertension prevalence between **smokers vs non-smokers**.

::: fragment
**A.** Independent t-test   **B.** Chi-square test   **C.** Pearson correlation   **D.** Linear regression
:::

::: fragment
**Answer: B.** Two categorical variables → chi-square (or Fisher's exact if any expected cell \< 5).
:::

# Module 6 — Introduction to Regression {background-color="#0B2545"}

## 🧭 Where are we? — ... → Bivariate ✅ → **Regression** → Interpretation

------------------------------------------------------------------------

### Why Regression? A Confounding Scenario

> Crude analysis: smokers have much higher SBP than non-smokers. But smokers in this sample are also, on average, **older**.

::: fragment
Is the SBP difference really about smoking — or partly about age?
:::

::: fragment
**Regression lets us ask: "What is the effect of smoking, holding age constant?"**
:::

------------------------------------------------------------------------

### What determines your choice of regression analysis?

::: fragment
**What variable type is your outcome variable of interest?**

Continuous -\> Linear regression

Dichotomous/Categorical -\> Logistic regression
:::

::: fragment
\*\* How many variables are present?

| Type of Analysis | Number of Dependent Variables (Y) (The Outcome/Goal) | Number of Independent Variables (X) (The Inputs/Predictors) | Simple Real-World Example |
|------------------|--------------------|------------------|------------------|
| Bivariate | 1 | 1 | You want to see if daily sodium intake (X) can predict a patient's blood pressure (Y) |
| Multivariate | 1 (usually) | 2 or more | You want to see how sodium intake (X1), age (X2), and weight (X3) collectively predict a patient's blood pressure (Y) |
:::

------------------------------------------------------------------------

### Bivariate → Confounding → Multivariable

```{mermaid}
flowchart LR
  Biv[Bivariate<br>Association] --> Conf{Potential<br>Confounding?}
  Conf -->|Yes| Multi[Multivariable<br>Analysis]
  Conf -->|No| Report[Report Crude<br>Estimate]
```

::: notes
Confounder = associated with both exposure and outcome, not on the causal pathway. Contrast briefly with mediator (on the pathway) and effect modifier (changes the effect size itself, not merely bias).
:::

------------------------------------------------------------------------

### Linear Regression — Continuous Outcome

**Example:** SBP predicted by age and BMI

```{r lm-demo}
model_lm <- lm(systolic_bp ~ age + bmi, data = hypertension_demo)
summary(model_lm)$coefficients |> round(3)
```

::: fragment
Each β = expected change in SBP per **1-unit increase**, holding the other variable constant.
:::

------------------------------------------------------------------------

### Reading the Linear Regression Output

| Term            | Meaning                                                 |
|-------------------------|-----------------------------------------------|
| β (coefficient) | Direction & magnitude of association                    |
| 95% CI          | Plausible range for the true β                          |
| p-value         | Compatibility with β = 0                                |
| R²              | Proportion of outcome variance explained (conceptually) |

```{r lm-plot}
#| fig-height: 3.2
ggplot(hypertension_demo, aes(age, systolic_bp)) +
  geom_point(alpha = 0.25, color = pal$teal) +
  geom_smooth(method = "lm", color = pal$navy) +
  theme_workshop() + labs(title = "SBP ~ Age (unadjusted view)")
```

------------------------------------------------------------------------

### Binary Logistic Regression — Binary Outcome

**Example:** Hypertension (Yes/No) predicted by age, BMI, smoking

```{r glm-demo}
model_glm <- glm(hypertension ~ age + bmi + smoking, family = binomial,
                  data = hypertension_demo)
exp(cbind(OR = coef(model_glm), confint(model_glm))) |> round(2)
```

::: notes
Point out: exponentiating the logit coefficient converts it to an Odds Ratio — this single transformation confuses many residents.
:::

------------------------------------------------------------------------

### Odds, Odds Ratios, COR vs AOR

::: incremental
- **Odds** = P(event) / P(no event) — not the same as probability
- **Crude OR (COR)** = unadjusted association from bivariate analysis
- **Adjusted OR (AOR)** = association after controlling for confounders
- Reference category must always be stated explicitly
:::

------------------------------------------------------------------------

### 🔮 Interpret Before We Reveal

::: r-fit-text
AOR = 2.1 (95% CI: 1.3–3.4)
:::

::: fragment
1.  Direction? → Positive association (AOR \> 1)
2.  Does CI include 1 (the null)? → No → statistically significant at α = 0.05
3.  Magnitude? → Exposed group has \~2.1× the odds of the outcome, adjusting for covariates
4.  Causation? → **No** — this alone does not establish causality
:::

------------------------------------------------------------------------

### Forest-Plot Style Illustration

```{r forest-demo}
#| fig-height: 3.5
forest_df <- tibble(
  variable = c("Age (per year)", "BMI (per unit)", "Smoking (vs non-smoker)"),
  OR = c(1.05, 1.08, 1.9),
  lower = c(1.02, 1.03, 1.2),
  upper = c(1.08, 1.14, 3.0)
)

ggplot(forest_df, aes(x = OR, y = fct_rev(variable))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = lower, xmax = upper), color = pal$navy, size = 0.8) +
  scale_x_log10() +
  theme_workshop() +
  labs(title = "Adjusted Odds Ratios (illustrative)", x = "AOR (log scale)", y = NULL)
```

------------------------------------------------------------------------

### Big Idea {.center background-color="#0B2545"}

::: r-fit-text
Regression **adjusts** estimates.

It does **not** rescue a poorly designed study.
:::

# Module 7 — Dissertation Pitfalls, Integration & Closing {background-color="#0B2545"}

## 🧭 Where are we? — ... → Regression ✅ → **Integration & Reporting**

------------------------------------------------------------------------

### Common Dissertation Pitfalls {.smaller}

::::: columns
::: {.column width="50%"}
- Objective ↔ test mismatch
- Unjustified categorization of continuous variables
- Incorrect denominators
- Reporting only p-values
- Missing effect estimates / CIs
:::

::: {.column width="50%"}
- Vague description of sampling
- Analyzing before cleaning
- Parametric vs non-parametric testing pitfall
- Regression variables with no conceptual rationale
- OR/AOR misinterpreted as relative risk
:::
:::::

------------------------------------------------------------------------

### 🔍 Spot the Problem — Mock Table

| Variable                     | p-value     |
|------------------------------|-------------|
| Age                          | 0.041237    |
| Smoking (n = 61/300, n = 45/442) | 0.002891029 |

::: fragment
Problems: only p-values (no estimates/CIs), excessive decimal places, inconsistent denominators for smoking (300 vs 442 — missing data unexplained).
:::

------------------------------------------------------------------------

### Five Rules to Take Home {background-color="#0B2545"}

::: r-fit-text
1.  Design analysis plan from the start (conception of project) - keep your research objective in mind   
2.  Know your variables   
3.  Clean and describe before testing   
4.  Report estimates and uncertainty — not p-values alone   
5.  Use regression thoughtfully, not automatically
:::

------------------------------------------------------------------------

### 🗳️ Closing Poll — Compare to Where We Started

**POLL:** *How confident are you now in choosing the correct statistical test for your dissertation objective?*

- A. Very confident
- B. Somewhat confident
- C. Not confident yet — and that's OK, here's your cheat sheet

------------------------------------------------------------------------

### Thank You {.center background-color="#0B2545"}

::: r-fit-text
Questions?
:::

Take-home cheat sheet, analysis plan template, and this dataset (simulated) are available in the workshop folder.

::: notes
Direct participants to `resources/analysis-cheatsheet.qmd` and `data/hypertension_demo.csv`. Remind them the dataset is entirely simulated for teaching purposes.
:::

